\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}

### 2025-04-21 paper/main.tex
Add attention scores

    \item \textbf{Attention-Based}: A more sophisticated approach using a finetuned model to extract attention scores across the dataset. For each token, attention patterns across all layers are aggregated to determine its contextual importance.

@@

\begin{table*}[ht!]
\centering
\small
\caption{Comparison of vocabulary pruning techniques on GLUE benchmark tasks. Results show accuracy on the development set with varying pruning methods. Best results for each task are highlighted in \textbf{bold}.}
\label{tab:results}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lccccccccc|c}
\toprule
\textbf{Method} & \textbf{Param. Red.} & \textbf{SST-2} & \textbf{MRPC} & \textbf{MNLI} & \textbf{CoLA} & \textbf{STS-B} & \textbf{QQP} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
\midrule
\multicolumn{11}{l}{\textit{Baseline}} \\
ModernBERT (Full) & 0.03\% & 0.951 & 0.856 & 0.881 & 0.586 & 0.917 & 0.917 & 0.916 & 0.598 & 0.828 \\
\midrule
\multicolumn{11}{l}{\textit{Vocabulary Pruning}} \\
Train Tokens Only & 6.7-22.6\% & 0.950 & 0.831 & \textbf{0.883} & 0.509 & \textbf{0.917} & \textbf{0.917} & 0.915 & 0.598 & 0.815 \\
\midrule
Random Selection & 9.0-23.3\% & 0.911 & 0.798 & 0.832 & 0.470 & 0.845 & 0.902 & 0.895 & 0.522 & 0.772 \\
Clustering-Based & 9.0-23.3\% & 0.899 & 0.714 & 0.712 & 0.000 & 0.786 & 0.875 & 0.836 & 0.510 & 0.667 \\
Frequency-Based & 9.0-23.3\% & 0.943 & 0.812 & 0.880 & 0.467 & 0.905 & 0.916 & \textbf{0.920} & 0.542 & 0.798 \\
TF-IDF Based & 9.0-23.3\% & 0.933 & 0.807 & 0.875 & 0.520 & 0.901 & 0.900 & 0.917 & \textbf{0.606} & \textbf{0.807} \\
\midrule
\multicolumn{11}{l}{\textit{Vocabulary Pruning with OOV Handling}} \\
Freq + OOV & 9.0-23.2\% & \textbf{0.938} & 0.828 & \textbf{0.881} & 0.540 & 0.906 & 0.916 & 0.918 & 0.538 & 0.808 \\
TF-IDF + OOV & 9.0-23.2\% & 0.923 & \textbf{0.834} & 0.878 & \textbf{0.615} & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table}[ht!]
\centering
\caption{Average performance change across GLUE tasks with different pruning methods.}
\label{tab:avg_performance}
\begin{tabular}{lcc}
\toprule
\textbf{Method} & \textbf{Avg. Param. Reduction} & \textbf{Avg. Performance Drop} \\
\midrule
Train Tokens Only & 14.22\% & 1.25\% \\
Random Selection & 16.44\% & 5.58\% \\
Clustering-Based & 16.44\% & 16.24\% \\
Frequency-Based & 16.44\% & 2.32\% \\
TF-IDF Based & 16.44\% & 2.04\% \\
Frequency + OOV & 16.40\% & 1.82\% \\
TF-IDF + OOV & 16.40\% & 0.96\% \\
\bottomrule
\end{tabular}
\end{table}

>>>>

    \item \textbf{Attention-Based}: This method uses attention patterns from fine-tuning to determine token importance. For each token, attention scores across all layers and heads are aggregated to capture its contextual relevance within the specific task. Tokens receiving consistently higher attention are considered more important for model decisions, providing a more nuanced importance measure that captures both frequency and semantic significance.

@@

\begin{table*}[ht!]
\centering
\small
\caption{Comparison of vocabulary pruning techniques on GLUE benchmark tasks. Results show accuracy on the development set with varying pruning methods. Best results for each task are highlighted in \textbf{bold}.}
\label{tab:results}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lccccccccc|c}
\toprule
\textbf{Method} & \textbf{Param. Red.} & \textbf{SST-2} & \textbf{MRPC} & \textbf{MNLI} & \textbf{CoLA} & \textbf{STS-B} & \textbf{QQP} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
\midrule
\multicolumn{11}{l}{\textit{Baseline}} \\
ModernBERT (Full) & 0.03\% & 0.951 & 0.856 & 0.881 & 0.586 & 0.917 & 0.917 & 0.916 & 0.598 & 0.828 \\
\midrule
\multicolumn{11}{l}{\textit{Vocabulary Pruning}} \\
Train Tokens Only & 6.7-22.6\% & 0.950 & 0.831 & 0.883 & 0.509 & 0.917 & 0.917 & 0.915 & 0.598 & 0.815 \\
\midrule
Random Selection & 9.0-23.3\% & 0.911 & 0.798 & 0.832 & 0.470 & 0.845 & 0.902 & 0.895 & 0.522 & 0.772 \\
Clustering-Based & 9.0-23.3\% & 0.899 & 0.714 & 0.712 & 0.000 & 0.786 & 0.875 & 0.836 & 0.510 & 0.667 \\
Frequency-Based & 9.0-23.3\% & 0.943 & 0.812 & 0.880 & 0.467 & \textbf{0.905} & \textbf{0.916} & \textbf{0.920} & 0.542 & 0.798 \\
Attention-Based & 8.8-23.1\% & \textbf{0.947} & \textbf{0.864} & 0.864 & \textbf{0.589} & 0.885 & 0.912 & 0.912 & 0.550 & 0.803 \\
TF-IDF Based & 9.0-23.3\% & 0.933 & 0.807 & \textbf{0.875} & 0.520 & 0.901 & 0.900 & 0.917 & \textbf{0.606} & \textbf{0.807} \\
\midrule
\multicolumn{11}{l}{\textit{Vocabulary Pruning with OOV Handling}} \\
Freq + OOV & 9.0-23.2\% & \textbf{0.938} & 0.828 & \textbf{0.881} & 0.540 & 0.906 & 0.916 & 0.918 & 0.538 & 0.808 \\
TF-IDF + OOV & 9.0-23.2\% & 0.923 & 0.834 & 0.878 & 0.615 & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table}[ht!]
\centering
\caption{Average performance change across GLUE tasks with different pruning methods.}
\label{tab:avg_performance}
\begin{tabular}{lcc}
\toprule
\textbf{Method} & \textbf{Avg. Param. Reduction} & \textbf{Avg. Performance Drop} \\
\midrule
Train Tokens Only & 14.22\% & 1.25\% \\
Random Selection & 16.44\% & 5.58\% \\
Clustering-Based & 16.44\% & 16.24\% \\
Frequency-Based & 16.44\% & 2.32\% \\
Attention-Based & 15.92\% & 1.99\% \\
TF-IDF Based & 16.44\% & 2.04\% \\
Frequency + OOV & 16.40\% & 1.82\% \\
TF-IDF + OOV & 16.40\% & 0.96\% \\
\bottomrule
\end{tabular}
\end{table}
